### 2010-07-25 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: freemonad fits on one line
Desc@ICFP: use \muide for tagged stuffs
Desc@ICFP: fix spacing, fix redundance.
Desc@ICFP: re-re-shaping of freemo

\newcommand{\muide}[2]{\D{\({\mu}^{\!+}\)}\!\!_{#1}\:#2}

@@

In Section~\ref{sec:desc-free-monad}, we have built a free monad
operation for simple descriptions. The process is similar in the
indexed world. Namely, given an indexed functor, we derive the indexed
functor coding its free monad: \note{pwm: Whoa there. Maybe we should
  say something about IMonads in general before we get to this point?}
%
\[\stk{
\begin{array}{ll}
\FreeIMonad{\_}{} : & _{\PI{\V{I}}{\Set}}
                     \PITEL{\V{R}}{\TagIDesc{\V{I}}} 
                     \PITEL{\V{X}}{\V{I} \To \Set}\To 
                      \TagIDesc{\V{I}}
\end{array} \\
\FreeIMonad{\pair{\V{E}}{\V{F}}{}}{\V{I}}{\V{R}} \mapsto \\ \;\;
    \pair{\pair{\ListCons{\SYMBDVar}{(\fst{\V{E}})}} 
               {\LAM{\V{i}}
                \pair{\DConst{(\V{R}\: \V{i})}}
                     {(\snd{\V{E}})\: \V{i}}{}}{}}
         {\V{F}}{}
}\]

Just as in the universe of descriptions, this construction comes with
an obvious \return and a substitution operation, the \bind. Its
definition is the following:
%
\[\stk{
\begin{array}{@{}ll}
\substI : & _{\PI{\V{I}}{\Set}}
            \PI{\V{X}, \V{Y}}{\V{I} \To \Set}
            \PITEL{\V{R}}{\TagIDesc{\V{I}}} \\
          & (\V{X} \DotTo 
             \SYMBIMu_{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}) \To 
            \SYMBIMu_{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{X}})} \DotTo
            \SYMBIMu_{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}
\end{array} \\
\substI\: \V{X}\: \V{Y}\: \V{R}\: \V{\sigma}\: \V{i}\: \V{t} \mapsto \\
\qquad    \cataI\: (\toIDesc{\FreeIMonad{\V{R}}{}{\V{X}}})\:
                      \SYMBIMu{(\FreeIMonad{\V{R}}{}{\V{Y}})}\:
                      (\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma})\:
                      \V{i}\:
                      \V{t} 
}\]
% 
Where  $\F{applyI}$ is defined as follow:
%
\[\stk{
\begin{array}{@{}ll}
\F{applyI} : & _{\PI{\V{I}}{\Set}}
            \PITEL{\V{R}}{\TagIDesc{\V{I}}}
            \PI{\V{X}, \V{Y}}{\V{I} \To \Set} \\
          & (\V{X} \DotTo \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}) \To \\
          & \idescop{\toIDesc{\FreeIMonad{\V{R}}
                                         {\V{I}}
                                         {\V{X}}}}
                    {\V{I}}
                    {\IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}} \DotTo 
            \IMu{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\SYMBDVar}{\V{x}}{}   & \V{\sigma}\: \V{i}\: \V{x}                   \\
\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\V{c}}{\V{ys}}{} & \Con{\pair{\V{c}}{\V{ys}}{}}
\end{array}
}\]

The fact that indexed descriptions are closed under substitution
is potentially of considerable utility, if we can exploit this fact:
\[
\idescop{\V{\sigma} \V{D}}{\V{J}}{\V{X}} 
    \mapsto 
        \idescop{\V{D}}
                {\V{I}}
                {\LAM{\V{i}}
                     {\idescop{\V{\sigma} \V{i}}
                              {\V{J}}
                              {\V{X}}}}
        \quad \mbox{where}\;\V{\sigma}:\V{I}\To\IDesc{\V{J}}
\]
By observing that a description can be decomposed via substitution, we
split its meaning into a superstructure of substructures, e.g. a
`database containing salaries', ready for traversal operations
preserving the former and targeting the latter. We shall now turn to
some examples of indexed descriptions, as well as a practical use of
free monad construction.

@@

This language has values, conditional
expression, addition and comparison. Informally, their types are:
%
\[
\begin{array}{l@{\::\:\:}l}
\SYMBEval            & \Val{\V{ty}} \To \V{ty} \\
\SYMBEcond           & \Ebool \To \V{ty} \To \V{ty} \To \V{ty}  \\ 
\end{array}
\qquad
\begin{array}{l@{\::\:\:}l}
\SYMBEplus           & \Enat \To \Enat \To \Enat                           \\
\SYMBEle             & \Enat \To \Enat \To \Ebool                          \\
\end{array}
\]
%
The function $\SYMBVal$ interprets object language types in the
host language, so that arguments to $\SYMBEval$ fit their
expected type.
\[\stk{
\SYMBVal : \Ty \To \Set \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\Val{\Enat}   & \Nat \\
\Val{\Ebool}  & \Bool
\end{array}
}\]
We take $\Nat$ and $\Bool$ to represent
natural numbers and Booleans in the host language, equipped with
addition $\plusHost$ and comparison $\leHost$.

We describe our syntax in two steps. First, we give the pattern
functor of the free monad, defining the operators of the
language. Second, we apply the free monad construction to obtain a
language of closed terms and a language of open terms. The operations
are described by a tagged indexed description, indexing over object
language types $\Ty$. We note that some constructors are always
available, namely $\SYMBEval$ and $\SYMBEcond$. On the other hand,
$\SYMBEplus$ and $\SYMBEle$ constructors are index-dependent, with
$\SYMBEplus$ available just when building a $\Enat$, $\SYMBEle$ just
for $\Ebool$. The code, below, reflects this intuition, with the first
component uniformly offering $\SYMBEval$ and $\SYMBEcond$, the second
selectively offering $\SYMBEplus$ or $\SYMBEle$.
%
%%% \begin{figure}
%
\[\stk{
\stk{
\HExprD : \TagIDesc{\Ty} \\
\HExprD \mapsto \sqr{ \HExprAD , \HExprID } \\
} \smallskip\\
\stk{
\HExprAD : \ATagIDesc{\Ty} \\
\HExprAD \mapsto \vtup{l}{
   {\vtup{r}{\SYMBEval\\ \SYMBEcond \,}} \red{,} \;
      \LAM{\V{ty}}
      \vtup{l@{\:}l}{
      \DProd{\DConst{(\Val{\V{ty}})}&}{\DConst{\Unit}} \\
      \DProd{\DProd{\DVar{\Ebool}}{\DProd{\DVar{\V{ty}}}{\DVar{\V{ty}}}}&}
        {\DConst{\Unit}} \\
     }
   }
\smallskip\\
} 
\\
\stk{
\HExprID : \ITagIDesc{\Ty} \\
\HExprID \mapsto \vtup{l}{
                   \vtup{r}{\sqr{\SYMBEplus} \\ \sqr{\SYMBEle}} \red{,} \;
  \LAM{\_} \sqr{\DProd{\DProd{\DVar{\Enat}}{\DVar{\Enat}}}{\DConst{\Unit}}}
                   }
}
}\]

@@

@@

However, so far, we are only able to define and manipulate
\emph{closed} terms. Adding variables, it is possible to build and
manipulate \emph{open} terms, that is, terms in a context. The context
is represented by a snoc-list of types:
%
\[
\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
\Ctxt           & \Set \\
\SYMBCtxtEmpty  & \Ctxt \\
\SYMBCtxtSnoc   & \Ctxt \To \Ty \To \Ctxt
\end{array}
\]
%
An environment realises the context, associating a value for each type:
%
\[
\stk{
\SYMBEnv : \Ctxt \To \Set \\
\begin{array}{@{}l@{\:\:\mapsto\:\:}l}
\Env{\CtxtEmpty}                & \Unit \\
\Env{(\CtxtSnoc{\V{G}}{\V{S}})} & \TIMES{\Env{\V{G}}}{\Val{\V{S}}}
\end{array}
}\]
%
In this setting, we define $\SYMBVar$ by:
%
\[\stk{
\Var{}{} : \Ctxt \To \Ty \To \Set \\
\begin{array}{@{}ll@{\:\:\mapsto\:\:}l}
\Var{\CtxtEmpty}{& \V{T}}                & 
    \Void \\
\Var{(\CtxtSnoc{\V{G}}{\V{S}})}{& \V{T}} & 
    \SUM{(\Var{\V{G}}{\V{T}})}{(\V{S} \PropEq \V{T})}
\end{array}
}\]
%
While $\SYMBVal$ maps the type to the corresponding host type,
$\SYMBVar$ indexes a value in the context, obtaining a proof that the
types match. The $\SYMBlookup$ function precisely follow this
semantics:
%
\[\stk{
\SYMBlookup : \PI{\V{G}}{\Ctxt} 
          \Env{\V{G}} \To 
          \PI{\V{T}}{\Ty} 
          \Var{\V{G}}{\V{T}} \To
          \Val{\V{T}} \\
\begin{array}{@{}l@{}l@{}l@{}l@{}lll}
\lookup{& (\CtxtSnoc{\V{G}}{.T})}{& \pair{\V{g}}{\V{t}}{}}{& \V{T}}{& (\SumRight{\C{refl}})} & \mapsto & \V{t} \\
\lookup{& (\CtxtSnoc{\V{G}}{\V{S}})}{& \pair{\V{g}}{\V{t}}{}}{& \V{T}}{& (\SumLeft{\V{x}})} & \mapsto & \lookup{\V{G}}{\V{g}}{\V{T}}{\V{x}} 
\end{array}
}\]

@@

Consequently, taking the free monad of \(\HExprD\) by \(\SYMBVar\:
\V{G}\), we obtain the language of open terms in a context \(\V{G}\):
%
\[
\openTerm{\V{G}} \mapsto \FreeIMonad{\HExprD}{\Ty}{(\SYMBVar\:\V{G})}
\]
%
The language of closed terms corresponds to the free monad assigning
an empty set of values to variables
%
\[
\closeTerm \mapsto \FreeIMonad{\HExprD}{\Ty}{\SYMBEmpty}
\]
%
With
%
\[\stk{
\SYMBEmpty : \Ty \To \Set \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\Empty{\_}   & \Zero \\
\end{array}
}\]
%

Given the syntax, let us supply the semantics of closed terms. We
implement an evaluator as a catamorphism:
%
\[\stk{
\evalH : \PI{\V{ty}}{\Ty} 
           \IMu{\Ty}{\closeTerm}{\V{ty}} \To
           \Val{\V{ty}} \\
\evalH\: \V{ty}\: \V{tm} \mapsto \cataI_{\Ty} \:
                                 (\toIDesc{\closeTerm})\: 
                                 \SYMBVal\: 
                                 \evalOne\: 
                                 \V{ty}\: 
                                 \V{tm}
}\]
%
To finish the job, we must supply the algebra which implements a single
step of evaluation, given subexpressions evaluated already.
%
\[\stk{
\evalOne : \PI{\V{ty}}{\Ty}
 \idescop{(\toIDesc{\closeTerm})\:\V{ty}}{\Ty}{\SYMBVal}
           \To {\Val{\V{ty}}} \\
\begin{array}{@{}l@{}c@{}l@{\:\mapsto\:\:}l}
\evalOne\: & \_\: & (\SYMBEval\;\V{x})                   & \V{x} \\
\evalOne\: & \_\: & (\SYMBEcond\:\BoolTrue\:\V{x}\:\_)   & \V{x} \\
\evalOne\: & \_\: & (\SYMBEcond\:\BoolFalse\:\_\:\V{y})  & \V{y} \\
\evalOne\: & \Enat\: & (\SYMBEplus\:\V{x}\:\V{y})   & \V{x} \plusHost \V{y} \\
\evalOne\: & \Ebool\: & (\SYMBEle\:\V{x}\:\V{y})  & \V{x} \leHost \V{y} 
\end{array}
}\]

Hence, we have a type-safe syntax and a tagless interpreter for our
closed term language, in the spirit
of~\citet{augustsson.carlsson:dependent.interpreter}, with help from
the generic catamorphism.

@@

The $\discharge$ function is relative to the required type and a
context of the right type. Its action is to map values to themself,
and variables to their value in context. This corresponds to the
following function:
%
\[\stk{
\begin{array}{@{}ll}
\discharge : & \PI{\V{G}}{\Ctxt}
               \Env{\V{G}} \To 
               \Var{\V{G}}{} \DotTo
               \IMu{\Ty}{\closeTerm}{}
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\discharge\: \V{G}\: \V{g}\: \V{ty}\: \V{v} &
    \Con{\pair{\SYMBEval}{\lookup{\V{G}}{\V{g}}{\V{ty}}{\V{v}}}{}}
\end{array}
}\]

@@

We are now left with applying $\discharge$ over all variables of the
term.  We simply have to fill in the right arguments to $\substI$, the
type guiding us:
%
\[
\stk{
\begin{array}{@{}ll}
\substH  : & \PI{\V{G}}{\Ctxt} \\
           & (\Var{\V{G}}{} \DotTo
              \IMu{\Ty}{\closeTerm}{}) \DotTo \\
          & \IMu{\Ty}{(\openTerm{\V{G}})}{} \DotTo 
            \IMu{\Ty}{\closeTerm}{}
\end{array} \\
\substH\: \V{G}\:
          \V{ty}\:          
          \V{g}\:
          \V{\sigma}\: 
          \V{tm} \mapsto  
\substI_{\Ty}\:
               (\SYMBVar\: \V{G})\: 
               \SYMBEmpty\:
               \HExprD\: 
               \V{\sigma}\:
               \V{ty}\:
               \V{tm}
}\]

Hence completing our implementation of the open terms interpreter.
We have defined a well-typed language of arithmetical expressions,
taking advantage of indexing. Then, we have implemented an evaluator
for closed term, based on the generic catamorphism function. Using the
free monad construction, we have automatically derived the language of
open terms. Using its monadic structure, we have implemented the
interpreter for open terms in context. Hence, without much effort, we
have described the syntax of a well-typed language, together with its
semantics.

>>>>

\newcommand{\SYMBmuide}{\D{\({\mu}^{\!+}\)}\xspace}
\newcommand{\muide}[2]{\SYMBmuide\!\!_{#1}\:#2}

@@

@@

This language has values, conditional expression, addition and
comparison. Informally, their types are:
%
\[
\begin{array}{l@{\::\:\:}l}
\SYMBEval            & \Val{\V{ty}} \To \V{ty} \\
\SYMBEcond           & \Ebool \To \V{ty} \To \V{ty} \To \V{ty}  \\ 
\end{array}
\qquad
\begin{array}{l@{\::\:\:}l}
\SYMBEplus           & \Enat \To \Enat \To \Enat                           \\
\SYMBEle             & \Enat \To \Enat \To \Ebool                          \\
\end{array}
\]
%
The function $\SYMBVal$ interprets object language types in the
host language, so that arguments to $\SYMBEval$ fit their
expected type.
%
\[\stk{
\SYMBVal : \Ty \To \Set \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\Val{\Enat}   & \Nat \\
\Val{\Ebool}  & \Bool
\end{array}
}\]
%
We take $\Nat$ and $\Bool$ to represent natural numbers and Booleans
in the host language, equipped with addition $\plusHost$ and
comparison $\leHost$.

We express our syntax as a tagged indexed description, indexing over
object language types $\Ty$. We note that some constructors are always
available, namely $\SYMBEval$ and $\SYMBEcond$. On the other hand,
$\SYMBEplus$ and $\SYMBEle$ constructors are index-dependent, with
$\SYMBEplus$ available just when building a $\Enat$, $\SYMBEle$ just
for $\Ebool$. The code, below, reflects this intuition, with the first
component uniformly offering $\SYMBEval$ and $\SYMBEcond$, the second
selectively offering $\SYMBEplus$ or $\SYMBEle$.
%
%%% \begin{figure}
%
\[\stk{
\stk{
\HExprD : \TagIDesc{\Ty} \\
\HExprD \mapsto \sqr{ \HExprAD , \HExprID } \\
} \smallskip\\
\stk{
\HExprAD : \ATagIDesc{\Ty} \\
\HExprAD \mapsto \vtup{l}{
   {\vtup{r}{\SYMBEval\\ \SYMBEcond \,}} \red{,} \;
      \LAM{\V{ty}}
      \vtup{l@{\:}l}{
      \DProd{\DConst{(\Val{\V{ty}})}&}{\DConst{\Unit}} \\
      \DProd{\DProd{\DVar{\Ebool}}{\DProd{\DVar{\V{ty}}}{\DVar{\V{ty}}}}&}
        {\DConst{\Unit}} \\
     }
   }
\smallskip\\
} 
\\
\stk{
\HExprID : \ITagIDesc{\Ty} \\
\HExprID \mapsto \vtup{l}{
                   \vtup{r}{\sqr{\SYMBEplus} \\ \sqr{\SYMBEle}} \red{,} \;
  \LAM{\_} \sqr{\DProd{\DProd{\DVar{\Enat}}{\DVar{\Enat}}}{\DConst{\Unit}}}
                   }
}
}\]

@@

Given the syntax, let us supply the semantics. We implement an
evaluator as a catamorphism:
%
\[\stk{
\evalH : \PI{\V{ty}}{\Ty} 
         \muide{\Ty}{\HExprD}\: \V{ty} \To
         \Val{\V{ty}} \\
\evalH\: \V{ty}\: \V{term} \mapsto \cataI_{\Ty} \:
                                 (\toIDesc{\HExprD})\: 
                                 \SYMBVal\: 
                                 \evalOne\: 
                                 \V{ty}\: 
                                 \V{term}
}\]
%
To finish the job, we must supply the algebra which implements a single
step of evaluation, given subexpressions evaluated already.
%
\[\stk{
\evalOne : \PI{\V{ty}}{\Ty}
 \idescop{(\toIDesc{\HExprD})\:\V{ty}}{\Ty}{\SYMBVal}
           \To {\Val{\V{ty}}} \\
\begin{array}{@{}l@{}c@{}l@{\:\mapsto\:\:}l}
\evalOne\: & \_\: & (\SYMBEval\;\V{x})                                            & \V{x} \\
\evalOne\: & \_\: & (\SYMBEcond\:\BoolTrue\:\V{x}\:\_)   & \V{x} \\
\evalOne\: & \_\: & (\SYMBEcond\:\BoolFalse\:\_\:\V{y})  & \V{y} \\
\evalOne\: & \Enat\: & (\SYMBEplus\:\V{x}\:\V{y})   & \V{x} \plusHost \V{y} \\
\evalOne\: & \Ebool\: & (\SYMBEle\:\V{x}\:\V{y})  & \V{x} \leHost \V{y} 
\end{array}
}\]

Hence, we have a type-safe syntax and a tagless interpreter for our
language, in the spirit
of~\citet{augustsson.carlsson:dependent.interpreter}, with help from
the generic catamorphism. However, so far, we are only able to define
and manipulate \emph{closed} terms. Adding variables, it is possible
to build and manipulate \emph{open} terms, that is, terms in a
context. We shall get this representation, for free, thanks to the
\emph{free indexed monad} construction.

In Section~\ref{sec:desc-free-monad}, we have built a free monad
operation for simple descriptions. The process is similar in the
indexed world. Namely, given an indexed functor, we derive the indexed
functor coding its free monad: \note{pwm: Whoa there. Maybe we should
  say something about IMonads in general before we get to this point?}
%
\[\stk{
\begin{array}{ll}
\FreeIMonad{\_}{} : & _{\PI{\V{I}}{\Set}}
                     \PITEL{\V{R}}{\TagIDesc{\V{I}}} 
                     \PITEL{\V{X}}{\V{I} \To \Set}\To 
                      \TagIDesc{\V{I}}
\end{array} \\
\FreeIMonad{\pair{\V{E}}{\V{F}}{}}{\V{I}}{\V{R}} \mapsto
    \pair{\pair{\ListCons{\SYMBDVar}{(\fst{\V{E}})}} 
               {\LAM{\V{i}}
                \pair{\DConst{(\V{R}\: \V{i})}}
                     {(\snd{\V{E}})\: \V{i}}{}}{}}
         {\V{F}}{}
}\]

Just as in the universe of descriptions, this construction comes with
an obvious \return and a substitution operation, the \bind. Its
definition is the following:
%
\[\stk{
\begin{array}{@{}ll}
\substI : & _{\PI{\V{I}}{\Set}}
            \PI{\V{X}, \V{Y}}{\V{I} \To \Set}
            \PITEL{\V{R}}{\TagIDesc{\V{I}}} \\
          & (\V{X} \DotTo 
             \muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}) \To 
             \muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{X}})} \DotTo
             \muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}
\end{array} \\
\substI\: \V{X}\: \V{Y}\: \V{R}\: \V{\sigma}\: \V{i}\: \V{t} \mapsto \\
\qquad    \cataI_{\V{I}}\: (\toIDesc{\FreeIMonad{\V{R}}{}{\V{X}}})\:
                      (\muide{\V{Y}}{(\FreeIMonad{\V{R}}{}{\V{Y}})})\:
                      (\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma})\:
                      \V{i}\:
                      \V{t} 
}\]
% 
Where  $\F{applyI}$ is defined as follow:
%
\[\stk{
\begin{array}{@{}ll}
\F{applyI} : & _{\PI{\V{I}}{\Set}}
            \PITEL{\V{R}}{\TagIDesc{\V{I}}}
            \PI{\V{X}, \V{Y}}{\V{I} \To \Set} \\
          & (\V{X} \DotTo \muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}) \To \\
          & \idescop{\toIDesc{\FreeIMonad{\V{R}}
                                         {\V{I}}
                                         {\V{X}}}}
                    {\V{I}}
                    {\muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}} \DotTo 
            \muide{\V{I}}{(\FreeIMonad{\V{R}}{\V{I}}{\V{Y}})}{}
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\SYMBDVar}{\V{x}}{}   & \V{\sigma}\: \V{i}\: \V{x}                   \\
\F{applyI}\: \V{R}\: \V{X}\: \V{Y}\: \V{\sigma}\: \V{i}\: \pair{\V{c}}{\V{ys}}{} & \Con{\pair{\V{c}}{\V{ys}}{}}
\end{array}
}\]

@@

In the previous section, we have presented a language of closed
arithmetic expressions. Using the free monad construction, we are
going to extend this construction to open terms. An open term is
defined with respect to a context, represented by a snoc-list of
types:
%
\[
\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
\Ctxt           & \Set \\
\SYMBCtxtEmpty  & \Ctxt \\
\SYMBCtxtSnoc   & \Ctxt \To \Ty \To \Ctxt
\end{array}
\]
%
An environment realises the context, associating a value for each type:
%
\[
\stk{
\SYMBEnv : \Ctxt \To \Set \\
\begin{array}{@{}l@{\:\:\mapsto\:\:}l}
\Env{\CtxtEmpty}                & \Unit \\
\Env{(\CtxtSnoc{\V{G}}{\V{S}})} & \TIMES{\Env{\V{G}}}{\Val{\V{S}}}
\end{array}
}\]
%
In this setting, we define $\SYMBVar$ by:
%
\[\stk{
\Var{}{} : \Ctxt \To \Ty \To \Set \\
\begin{array}{@{}ll@{\:\:\mapsto\:\:}l}
\Var{\CtxtEmpty}{& \V{T}}                & 
    \Void \\
\Var{(\CtxtSnoc{\V{G}}{\V{S}})}{& \V{T}} & 
    \SUM{(\Var{\V{G}}{\V{T}})}{(\V{S} \PropEq \V{T})}
\end{array}
}\]
%
While $\SYMBVal$ maps the type to the corresponding host type,
$\SYMBVar$ indexes a value in the context, obtaining a proof that the
types match. The $\SYMBlookup$ function precisely follow this
semantics:
%
\[\stk{
\SYMBlookup : \PI{\V{G}}{\Ctxt} 
          \Env{\V{G}} \To 
          \PI{\V{T}}{\Ty} 
          \Var{\V{G}}{\V{T}} \To
          \Val{\V{T}} \\
\begin{array}{@{}l@{}l@{}l@{}l@{}lll}
\lookup{& (\CtxtSnoc{\V{G}}{.T})}{& \pair{\V{g}}{\V{t}}{}}{& \V{T}}{& (\SumRight{\C{refl}})} & \mapsto & \V{t} \\
\lookup{& (\CtxtSnoc{\V{G}}{\V{S}})}{& \pair{\V{g}}{\V{t}}{}}{& \V{T}}{& (\SumLeft{\V{x}})} & \mapsto & \lookup{\V{G}}{\V{g}}{\V{T}}{\V{x}} 
\end{array}
}\]

@@

\newcommand{\update}{\F{update}}

Consequently, taking the free monad of \(\HExprD\) by \(\SYMBVar\:
\V{G}\), we obtain the language of open terms in a context \(\V{G}\):
%
\[
\openTerm{\V{G}} \mapsto \FreeIMonad{\HExprD}{\Ty}{(\SYMBVar\:\V{G})}
\]
%
In this setting, the language of closed terms corresponds to the free
monad assigning an empty set of values to variables
%
\[
\closeTerm \mapsto \FreeIMonad{\HExprD}{\Ty}{\SYMBEmpty}
\]
%
With
%
\[\stk{
\SYMBEmpty : \Ty \To \Set \\
\begin{array}{@{}l@{\:\:\mapsto\:\:}l}
\Empty{\_}   & \Zero \\
\end{array}
}\]
%
Obviously, \(\closeTerm\) and \(\HExprD\) describe isomorphic
datatypes. For instance, you can update your old \(\HExprD\) to a
shiny \(\closeTerm\) with \(\update\):
%
\[\stk{
\update : \muide{\Ty}{\HExprD} \DotTo \muide{\Ty}{\closeTerm} \\
\begin{array}{@{}l@{}l}
\update\: \V{ty}\: \V{tm} \mapsto \cataI_{\Ty}\: & \Ty\:
                                                  (\toIDesc{\HExprD})\:
                                                  (\muide{\Ty}{\closeTerm})\: \\
                                                & (\LAM{\_} \LAM{\pair{\V{tag}}{\V{tm}}{}} \Con{\pair{\Su{\V{tag}}}{\V{tm}}{}})\:
                                                  \V{ty}\:
                                                  \V{tm}
\end{array}
}\]
% 
The other direction of the isomorphism is straightforward, the
$\SYMBDVar$ case being impossible. Therefore, we are entitled to
reuse the $\evalH$ function to define the semantics of $\closeTerm$.

@@

The $\discharge$ function is relative to the required type and a
context of the right type. Its action is to map values to themself,
and variables to their value in context. This corresponds to the
following function:
%
\[\stk{
\begin{array}{@{}ll}
\discharge : & \PI{\V{G}}{\Ctxt}
               \Env{\V{G}} \To 
               \Var{\V{G}}{} \DotTo
               \muide{\Ty}{\closeTerm}
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\discharge\: \V{G}\: \V{g}\: \V{ty}\: \V{v} &
    \Con{\pair{\SYMBEval}{\lookup{\V{G}}{\V{g}}{\V{ty}}{\V{v}}}{}}
\end{array}
}\]

@@

We are now left with applying $\discharge$ over all variables of the
term.  We simply have to fill in the right arguments to $\substI$, the
type guiding us:
%
\[
\stk{
\begin{array}{@{}ll}
\substH  : & \PI{\V{G}}{\Ctxt} \\
           & (\Var{\V{G}}{} \DotTo
              \muide{\Ty}{\closeTerm}) \DotTo \\
          & \muide{\Ty}{(\openTerm{\V{G}})} \DotTo 
            \muide{\Ty}{\closeTerm}
\end{array} \\
\substH\: \V{G}\:
          \V{ty}\:          
          \V{g}\:
          \V{\sigma}\: 
          \V{tm} \mapsto  
\substI_{\Ty}\:
               (\SYMBVar\: \V{G})\: 
               \SYMBEmpty\:
               \HExprD\: 
               \V{\sigma}\:
               \V{ty}\:
               \V{tm}
}\]

Hence completing our implementation of the open terms interpreter.
We have defined a well-typed language of arithmetical expressions,
taking advantage of indexing. Then, we have implemented an evaluator
for closed term, based on the generic catamorphism function. Using the
free monad construction, we have automatically derived the language of
open terms. Using its monadic structure, we have implemented the
interpreter for open terms in context. Without much effort, we
have described the syntax of a well-typed language, together with its
semantics.

The fact that indexed descriptions are closed under substitution
is potentially of considerable utility, if we can exploit this fact:
\[
\idescop{\V{\sigma} \V{D}}{\V{J}}{\V{X}} 
    \mapsto 
        \idescop{\V{D}}
                {\V{I}}
                {\LAM{\V{i}}
                     {\idescop{\V{\sigma} \V{i}}
                              {\V{J}}
                              {\V{X}}}}
        \quad \mbox{where}\;\V{\sigma}:\V{I}\To\IDesc{\V{J}}
\]
By observing that a description can be decomposed via substitution, we
split its meaning into a superstructure of substructures, e.g. a
`database containing salaries', ready for traversal operations
preserving the former and targeting the latter. We shall now turn to
some practical use of the free monad construction.
